Exploring the composition of lithic assemblages in Mesolithic south-eastern Norway

### 2021-07-29 analysis/paper/paper.Rmd
Cutting down text

---
title: "Exploring the composition of lithic assemblages in Mesolithic south-eastern Norway"
header-includes:
   - \usepackage{lineno}
   - \linenumbers
author:
  - Isak Roalkvam:
      email: isak.roalkvam@iakh.uio.no
      institute: [UiO]
      correspondence: true
institute:
  - UiO: University of Oslo, Department of Archaeology, History and Conservation
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      toc: false
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/jas_reports.csl" # Insert path for the bib-style
abstract: |
  This paper leverages multivariate statistics to explore the composition of 54 Mesolithic assemblages located in south-eastern Norway. To provide analytical control pertaining to factors such as variable excavation practices, systems for artefact categorisation and raw-material availability, the sites chosen for analysis have all been excavated relatively recently and have a constrained geographical distribution. The assemblages were explored following two strains of analysis. The first of these entailed the use of artefact categories that are established within Norwegian Mesolithic archaeology, while the other involved drawing on measures that have been linked directly to land-use and mobility patterns associated with lithic assemblages more widely. The findings pertaining to the established artefact categories largely reflect the temporal development previously reported in Norwegian Mesolithic research, which has been based on more subjectively driven methods. Furthermore, the chronological trends associated with variables taken from the so-called Whole Assemblage Behavioural Index, originally developed for characterising Palaeolithic assemblages in terms of associated mobility patterns, also align with the development previously proposed in the literature. This provides an initial indication that these measures are applicable in a Norwegian Mesolithic setting as well, setting the stage for a more targeted and rigorous model evaluation outside this exploratory setting. This might ultimately yield a powerful comparative tool for more extensive analyses of Mesolithic assemblages.

@@

Even though each individual assemblage can have been impacted by an virtual infinitude of effects that might skew an archaeological interpretation, this does not preclude the applicability of inductive analyses aimed at revealing overarching structure in the data without imposing overly complex analytical frameworks that attempt to account for these particularities [@bevan2015]. Structure that can be revealed from considering all of the assemblages in aggregate can constitute a step in an iterative analytical chain that ultimately aims to tease apart the multitude of factors that have shaped the composition of the assemblages, and should be of value to subsequent in-depth studies of any individual site. The most immediate danger of the approach outlined here is rather to be overly naive in the causal significance and cultural importance that is ascribed to any identified pattern. As such, the main aim of this analysis is to compare the results with findings reported in previous literature concerned with the Mesolithic in southern Norway and have the generation of new hypotheses as a possible outcome. To this end, the analysis follows two analytical avenues. The first involves an analysis of the assemblages using the classification of the artefacts done for the original excavation reports. The second involves an analysis of the assemblages in light of the so-called Whole Assemblage Behavioural Index [e.g. @clark2017], which has been employed in other contexts to align properties of lithic assemblages with land-use and mobility patterns.

@@

The Early Mesolithic, or Flake Axe Phase, is defined as lasting from c. 9200--8200 BCE [@reitan2016]. The phase is set to start with the first human occupation in Norway which is widely held as originating from southern Scandinavian and northern European regions. This is to be directly reflected by similarities in the artefact inventories [@bjerck2008; @bang-andersen2012; @berg-hansen2018; @fuglestvedt2012; @glørstad2016]. Previous research has typically proposed that the Early Mesolithic is characterised by a relatively high degree of mobility, and low variation in site types and associated mobility patterns [e.g. @bjerck2008; @breivik2016; @fuglestvedt2012; but see @viken2018]. The transition to the subsequent Middle Mesolithic, or Microlith Phase, at around 8200 BCE has been linked to changes in blade [@damlien2016] and subsequently axe technology [@eymundsson2018; @solheim2020], which in turn has been associated with changes in population genomics and related migration events hailing from the Eurasian steppes [@günther2018; @manninen2021]. The radiocarbon record points towards a coinciding population decline in southern Norway around this time [@nielsen2021]. Although this does not appear to be evident in the regional data for south-eastern Norway, taphonomic loss associated with these early dates is an issue [@solheim2018a; @solheim2020a; @nielsen2021]. In the chronological framework of @reitan2016, the Microlith phase is defined as lasting until around 7000 BCE. The Microlith phase is followed by the Pecked Adze Phase, characterised by a more dominating presence of non-flint macro tools and associated production waste in the assemblages [@reitan2016]. Following @reitan2016, the next typological transition at c. 5600 BCE signifies the onset of the Nøstvet Adze Phase. While previously defined as having a slightly longer duration, the Nøstvet Phase has traditionally been seen as representing the onset of more varied settlement systems and stable mobility patterns in the Mesolithic of south-eastern Norway [e.g. @jaksland2001; @lindblom1984] In recent years it has been suggested that the transition to a decrease in mobility and more varied land-use patterns can be traced back to the Middle Mesolithic [@solheim2016]. The subsequent Transverse Arrowhead Phase (c. 4500--3900 BCE) is characterised by a dramatic decrease in axe finds, and the introduction of tranverse-, tanged- and single-edged points [@reitan2016]. It has recently been suggested that a dispersal of people from southern Scandinavia into southern Norway takes place in this period [@eigeland2015, 379; @nielsen2021], which could follow after a preceding population decline at c. 4300 BCE [@nielsen2021].

The `r nrow(sites)` coastal sites chosen for analysis here have a relatively limited geographical distribution in south-eastern Norway (Figure \@ref(fig:dist)A). The sites were excavated as part of four larger excavation projects that all took place within the last 15 years [@solheim2013b; @jaksland2014a; @melvold2014b; @reitan2014l; @solheim2017b]. The sites included in the analysis consist of all Mesolithic sites excavated in conjunction with the projects that have assemblages holding more than 100 artefacts. The institution responsible for these excavations was the Museum of Cultural History in Oslo. This has led to a considerable overlap in the archaeological personnel involved, and comparable excavation practices across the excavations. Furthermore, with these projects, major efforts were made to standardise how lithic artefacts were to be classified at the museum [@melvold2014c; @koxvold2017]. As a result, this should reduce the amount of artificial patterning in the data incurred by discrepancies in the employed systems for categorisation [e.g. @clark2006; @dibble2017]. In this setting, for example, bias could potentially follow from the fact that two of the projects have sites with relatively contemporaneous dates [@jaksland2014; @solheim2013b, see also Figure \@ref(fig:dist)B]. Any project-dependent classification practice could as a consequence lead to an exaggeration of chronological differences between the assemblages. While this is difficult to fully account for, I do believe that the relative contemporaneity of the excavation projects, as well as the overlap in excavation and classification practices should minimise the above-mentioned effects.

@@

A defining characteristic of the Norwegian Mesolithic is that a clear majority of the known sites are located in coastal areas [e.g. @bjerck2008]. Furthermore, these coastal sites appear to predominantly have been located on or close to the contemporary shoreline when they were in use [e.g. @åstveit2018; @breivik2018; @møller1987; @solheim2020]. In south-eastern Norway, this pattern is combined with a continuous regression of the shoreline, following from isostatic rebound [e.g. @romundset2018; @sørensen1979]. The fairly rapid shoreline displacement means that the sites tend not to have retained their strategic or ecologically beneficial shore-bound location for long periods of time [cf. @perreault2019, 47]. Consequently, the shore-bound settlement, combined with the rapid shoreline displacement has resulted in a relatively high degree of spatial separation of cumulative palimpsests, to follow the terminology of @bailey2007, while the reconstruction of the trajectory of relative sea-level change allows for a relatively good control of when these accumulation events occurred. In other parts of the world, a higher degree of spatial distribution means that while the physical separation of material can help delineate discrete events, this typically comes at the cost of losing temporal resolution as any stratigraphic relationship between the events is lost [@bailey2007]. However, as the rate of isostatic rebound has varied throughout the Mesolithic in the region, and local topography and bathymetry will have impacted how rapidly a site lost its shore-bound location, this effect is not evenly distributed in time and space. In the earliest part of the Mesolithic, the displacement rate within the study area would have been around as much as 8.8 cm/year, falling to around 0.5 cm/year towards the end of the Mesolithic [@sørensen2014]. Thus, while relative sea-level change appears to have reduced the degree of mixing that has occurred in the assemblages, it is worth bearing in mind that this could vary depending on when and where they were in use, potentially reducing the degree to which their composition can be directly compared.

The lithic data analysed here is based on the classification of the site assemblages done for the original excavation reports, and consists of `r ncol(select(sites, core:flake_retouch_other))` variables representing different debitage and tool types. While the classification practices for the excavation projects were standardised to an extent, there are some instances where time was allocated to identify additional artefact sub-categories aimed at answering specific research questions. Some categories in the original reports have therefore been combined in the dataset. For example, this pertains to the category narrow-blades (width \> 8mm and \< 12mm), which was only separated from (macro-)blades (width $\ge$ 12 mm) and micro-blades (width $\le$ 8 mm) for some of the sites. Narrow-blades were combined with the blade category here. Furthermore, the artefact data have here been divided into flint and non-flint materials. Flint does not occur indigenously in Norway, and is only available locally as nodules that have been transported and deposited by retreating and drifting ice [e.g. @berg-hansen1999]. This means that the distribution and quality of flint has been impacted by a diverse set of factors relating to climatic and geographical factors such as, but not limited to, topographic variability, shoreline morphology and ocean currents [@eigeland2015, 46]. Thus, while flint is treated as a unified category here, the variability in quality could have been substantial [@eigeland2015, 45-53]. Furthermore, the various non-flint raw materials that have been lumped together have quite disparate properties, where fine-grained cryptocrystalline materials are often used as a substitute or supplement to flint, while other, coarser materials are usually associated with the production of axes and other macro tools. Given this differentiated use, these raw-material properties are expected to be reflected in the retained debitage and tool categories. An important benefit of combining all of the non-flint materials is that this reduces the dependency on whether or not these have been correctly and consistently categorised for the reports [cf. @frivoll2017]. While certainly a topic deserving of more attention, the general sentiment in the literature is that there would have been stable access to locally available non-flint raw-materials of good quality in south-eastern Norway [e.g. @eigeland2015, 370; @glørstad2011]. Finally, while factors such as landscape changes through shoreline displacement can have led to variable raw-material availability at the analysed sites, their relatively constrained geographical distribution hopefully counteracts some non-behavioural sources of variation.

@@

The aim of the first of part of the analysis is to evaluate the degree to which the composition of the assemblages align with patterns that have been suggested by earlier studies of the Mesolithic in southern Norway that have employed more informally driven methods. This consequently assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviourally meaningful. However, the approach taken is also partially informed by the so-called Frison effect [@jelinek1976], which pertains to the fact that lithics studied by archaeologists can have had long and complex use-lives in which they took on a multitude of different shapes before they were ultimately discarded. Several scholars have built on this to argue that morphological variation in retouched lithics from the Palaeolithic cannot be assumed to predominantly be the result of the intention of the original knapper to reach some desired end-product, but rather that what is commonly categorised as discrete types of artefacts by archaeologists can instead in large part be related to variable degrees of modification through use and rejuvenation [e.g. @barton1991; @dibble1995; @barton2021].

However, for the most part we lack even a most basic understanding of what any individual lithic object in an assemblage has been used for [@dibble2017]. For example, a vast amount of artefacts defined as debitage are likely to have fulfilled the function of tools, and both debitage and formal tool types could have had various different purposes and had a multitude of shapes throughout their use-life. While use-wear analysis could potentially offer a way to identify what artefacts were used for towards the end of their use-life, these kinds of analyses are extremely time-consuming and are therefore typically only conducted on a smaller number of artefacts that have already been selected for analysis based on their shape [e.g. @solheim2018]. Thus, while these analyses can potentially get at in-group variation pertaining to the end-state of a group of artefacts, they do not tell us whether or not their classification as a unified group is meaningful in the first place [@dibble2017]. As a consequence, the second part of the analysis employs a suite of measures developed for the classification of lithic assemblages with these inferential limitations in mind [see @barton2011; @clark2017, and below]. The logic behind these measures are founded on an understanding of technology as being organised along a continuum ranging between curated and expedient [@binford1973; @binford1977; @binford1979]. An expedient technological organisation pertains to the situational production of tools to meet immediate needs, with little investment of time and resources in modification and rejuvination, resulting in high rates of tool replacement. Curated technological organisation, on the other hand, has been defined as related to manufacture and maintenance of tools in anticipation of future use, the transport of these artefacts between places of use, and the modification and rejuvenation of artefacts for different and changing situations.

However, following not least from the ambiguous definition first put forward by @binford1973, the theoretical definition of curation, its archaeological correlates, and behavioural implications have been widely discussed and disputed, and no single definition has ever been reached [e.g. @bamforth1986; @nash1996; @shott1996; @surovell2009, 9--13]. The continuum between curated and expedient technology has for example been related to dimensions such as land-use and mobility strategies and raw-material quality and availability [e.g. @andrefsky1994; @parry1987; @kuhn1992; @smith2015; @clark2017]. Still, that the distinction can offer a useful analytical point of departure if clearly and explicitly operationalised seems more or less agreed upon, and some dimensions of the concept are generally accepted. For example, although precisely how it is measured may vary, the empirical correspondent to a curated technological organisation is typically defined by high degrees of retouch, as this is commonly seen as a means of realising the potential utility of a tool --- or extending its use-life --- by the repeated rejuvenation and modification of edges [e.g. @bamforth1986; @dibble1995; @shott2005].

Furthermore, one concrete operationalisation of the terms have been forwarded by @barton1998 and collegues [e.g. @barton2013; @barton1999; @riel-salvatore2004; @riel-salvatore2007; @barton2011; @barton2014; @clark2017; @villaverde1998], who through a series of studies have shown that the relationship between volumetric density of lithics and relative frequency of retouched artefacts in lithic assemblages have a consistent negative relationship across a wide range of chronological and cultural context, ranging from Pleistocene and Holocene assemblages in Europe and Asia, to assemblages associated with both Neanderthals and modern humans [@barton2011; @riel-salvatore2008]. This relationship is taken to reflect degree of curation, and is in turn mainly to follow from the accumulated nature of land-use and mobility patterns associated with the assemblages [@barton2014].

In this model, higher degree of mobility would mean a higher dependency on the artefacts and the material people could bring with them, and dimensions such as weight, reliability, repairability, and the degree to which artefacts could be manipulated to fulfil a wide range of tasks are therefore assumed to have been factors of concern. From this it follows that the empirical expectation for short-term camps is a curated technological organisation with higher relative frequency of retouched artefacts, and a lower overall density of lithics [@clark2017]. More time spent in a single location, on the other hand, is assumed to lead to better control of raw-material availability and to allow for its accumulation. This should in turn lead to a more expedient technological organisation with reduced necessity for the conservation of lithics and extensive use of retouch. The empirical expectation for lower degree of mobility is therefore relatively high density of lithics, a low relative frequency of retouched artefacts, as well as a higher number of cores and unretouched flakes and blades. These variables and underlying logic constitute what has been termed the whole assemblage behavioural index [WABI, @clark2017], and is the main framework adopted here.

However, as these measures are argued to predominantly be determined by land-use and mobility patterns, relative frequency of chips and relative frequency of non-flint material are also included in the analysis, as these measures have also been linked to mobility patterns and is of central importance in Norwegian Stone Age archaeology [e.g. @Bicho2020; @breivik2016a; @reitan2016; @kitchel2021]---the use of local non-flint material has been taken to indicate reduced mobility and increased familiarity with local surroundings [@jaksland2001, 112; @glørstad2010, 181].

This relationship has been related to the continuum defined by @binford1980 between residentially mobile foragers and logistically mobile collectors. Residential mobility involves the movement of entire groups between resource patches throughout the year, while logistic mobility entails the use of central base-camps from where task groups venture to retrieve resources, which can lead to a wider spectrum of site types and associated mobility patterns. While this framwork can offer

@@

The relatively constrained geographical distribution of the analysed sites, the limited temporal range over which they where investigated, as well as the methodological equivalency across excavation projects hopefully leads to an exclusion of some biases that might otherwise skew an exploratory analysis, rendering it more likely that behaviourally meaningful patterns are identified. However, the exploratory perspective means that a wide range of combinations and transformations of variables has been explored to identify patterning in the data. While only parts of this process can sensibly be reported here, the data and employed R programming script is freely available as a research compendium on GitHub (<https://github.com/isakro/exploring-assemblages-se-norway>) and Zenodo ([placeholder]), following @marwick2018b, allowing readers to explore and scrutinise the data and the final analytical choices made [@Marwick2017]. However, this inductive data-dredging or pattern-searching approach does constitute a limited inferential framework [@clark2009], as it involves a *post hoc* accommodation of explanations to meet the observed data---data that is both selectively and subjectively reported upon. The process can still provide the identification of empirical patterns with respects to the employed units of analysis, which in turn can form the basis for social and behavioural hypotheses. This can lay the foundation for a deductive research agenda with targeted model evaluation for which clear test implications can be derived [@clark2009, 29].

The `r nrow(sites)` analysed sites have been dated by reference to relative sea-level change, typology and/or radiocarbon dates (Table 1). Date ranges for sites based on shoreline displacement and typology are taken from the original reports and follow the evaluation done by the original excavators. Where radiocarbon age determinations believed to be associated with the lithic material are available, these have been calibrated using the IntCal20 calibration curve [@reimer2020] and subjected to Bayesian modelling using OxCal v4.4.4 [@bronkramsey2009] through the oxcAAR package [@hinz2021] for R [@rcoreteam]. The only constraint imposed for the modelling of the dates was that the dates from each site are assumed to represent a related group of events through the application of the Boundary function [@bronkramsey2021]. The resulting posterior density estimates were then summed for each site. Radiocarbon data is provided in the supplementary data, and has also been collated and reported by @solheim2020a.

The first part of the analysis of the lithic data involves employing the method of correspondence analysis (CA), using the lithic count data as classified for the original excavation reports. As this part of the analysis partially draws on the above-mentioned Frison effect, several artefact categories have been collapsed for the CA. A full overview of the aggregated variables and their constituent parts is provided in the supplementary material. An underlying assumption is therefore effectively that the retained categories represent artefact categories that have fulfilled different purposes or are related to different technological processes. While aggregating artefact categories in this manner could potentially subsume important variation, it does also reduce the possibility that any conclusions are not simply the result of employing erroneous units of analysis.

In sum, the variables employed in the analysis are relative frequency of secondarily worked lithics (RFSL), defined as the number of retouched or ground lithics divided by the assemblage total; volumetric density of lithics (VDL), defined as number of artefacts per excavated m³; relative frequency of chips, defined as the proportion of artefacts with size \< 0.1 cm; relative frequency of cores, simply the proportion of all artefacts classified as cores in the original reports; relative frequency of blanks, here defined as the proportion of all artefacts classified as flakes, blades, micro-blades or fragments; and finally relative frequency of non-flint material. Following @Bicho2020, the analysis is done using principal components analysis (PCA), leading to a shift in focus from the relative composition emphasised by the CA, to having more weight placed on patterning in the most abundant occurrences [@baxter1994, 71--77, 103].

>>>>

---
title: "Exploring the composition of lithic assemblages in Mesolithic south-eastern Norway"
header-includes:
   - \usepackage{lineno}
   - \linenumbers
author:
  - Isak Roalkvam:
      email: isak.roalkvam@iakh.uio.no
      institute: [UiO]
      correspondence: true
institute:
  - UiO: University of Oslo, Department of Archaeology, History and Conservation
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      toc: false
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/saa.csl" # Insert path for the bib-style
abstract: |
  This paper leverages multivariate statistics to explore the composition of 54 Mesolithic assemblages located in south-eastern Norway. To provide analytical control pertaining to factors such as variable excavation practices, systems for artefact categorisation and raw-material availability, the sites chosen for analysis have all been excavated relatively recently and have a constrained geographical distribution. The assemblages were explored following two strains of analysis. The first of these entailed the use of artefact categories that are established within Norwegian Mesolithic archaeology, while the other involved drawing on measures that have been linked directly to land-use and mobility patterns associated with lithic assemblages more widely. The findings pertaining to the established artefact categories largely reflect the temporal development previously reported in Norwegian Mesolithic research, which has been based on more subjectively driven methods. Furthermore, the chronological trends associated with variables taken from the so-called Whole Assemblage Behavioural Indicators (e.g. Clark and Barton 2017), originally devised for characterising Palaeolithic assemblages in terms of associated mobility patterns, also align with the development previously proposed in the literature. This provides an initial indication that these measures are applicable in a Norwegian Mesolithic setting as well, setting the stage for a more targeted and rigorous model evaluation outside this exploratory setting. Furthermore, this supports the notion that these measures can offer a powerful comparative tool in the analysis of lithic assemblages more generally.

@@

Even though each individual assemblage can have been impacted by an virtual infinitude of effects that might skew an archaeological interpretation, this does not preclude the applicability of inductive analyses aimed at revealing overarching structure in the data without imposing overly complex analytical frameworks that attempt to account for these particularities [@bevan2015]. Structure that can be revealed from considering all of the assemblages in aggregate can constitute a step in an iterative analytical chain that ultimately aims to tease apart the multitude of factors that have shaped the composition of the assemblages, and should be of value to subsequent in-depth studies of any individual site. The most immediate danger of the approach outlined here is rather to be overly naive in the causal significance and cultural importance that is ascribed to any identified pattern. As such, the main aim of this analysis is to compare the results with findings reported in previous literature concerned with the Mesolithic in southern Norway and have the generation of new hypotheses as a possible outcome. To this end, the analysis follows two analytical avenues. The first involves an analysis of the assemblages using the classification of the artefacts done for the original excavation reports. The second involves an analysis of the assemblages in light of the so-called Whole Assemblage Behavioural Indicators [e.g. @clark2017], which have been employed in other contexts to align properties of lithic assemblages with land-use and mobility patterns.

@@

The Early Mesolithic, or Flake Axe Phase, is defined as lasting from c. 9200--8200 BCE [@reitan2016] and is set to start with the first recorded human presence in Norway. Previous research has typically proposed that the Early Mesolithic is characterised by a relatively high degree of mobility, and low variation in site types and associated mobility patterns [e.g. @bjerck2008; @breivik2016; @fuglestvedt2012; but see @viken2018c]. Around the transition to the subsequent Middle Mesolithic, or Microlith Phase, at 8200 BCE pervasive changes in blade and axe technology takaes place [@damlien2016; @eymundsson2018; @solheim2020], which in turn has been associated with changes in population genomics and related migration events hailing from the Eurasian steppes [@günther2018; @manninen2021]. The Microlith Phase is defined as lasting until around 7000 BCE, which is followed by the Pecked Adze Phase, characterised by a more dominating presence of non-flint macro tools and associated production waste in the assemblages [@reitan2016]. The next typological transition at c. 5600 BCE signifies the onset of the Nøstvet Adze Phase. While previously defined as having a slightly longer duration, the Nøstvet Phase has traditionally been seen as representing the onset of more varied settlement systems and stable mobility patterns [e.g. @jaksland2001; @lindblom1984]. In recent years it has been suggested that the transition to a decrease in mobility and more varied land-use patterns can be traced back to the Middle Mesolithic [@solheim2016]. The subsequent Transverse Arrowhead Phase (c. 4500--3900 BCE) is characterised by a dramatic decrease in axe finds, and the introduction of new flint projectiles [@reitan2016]. It has recently been suggested that a dispersal of people from southern Scandinavia into southern Norway takes place in this period [@eigeland2015, 379; @nielsen2021], which could follow after a preceding population decline at c. 4300 BCE [@nielsen2021].

@@

A defining characteristic of the Norwegian Mesolithic is that a clear majority of the known sites are located in coastal areas [e.g. @bjerck2008]. Furthermore, these coastal sites appear to predominantly have been located on or close to the contemporary shoreline when they were in use [@åstveit2018; @breivik2018; @møller1987; @solheim2020]. In south-eastern Norway, this pattern is combined with a continuous regression of the shoreline, following from isostatic rebound [e.g. @romundset2018; @sørensen1979]. The fairly rapid shoreline displacement means that the sites tend not to have retained their strategic or ecologically beneficial shore-bound location for long periods of time [cf. @perreault2019, 47]. Consequently, the shore-bound settlement, combined with the rapid shoreline displacement has resulted in a relatively high degree of spatial separation of cumulative palimpsests, to follow the terminology of @bailey2007, while the reconstruction of the trajectory of relative sea-level change allows for a relatively good control of when these accumulation events occurred. In other parts of the world, a higher degree of spatial distribution means that while the physical separation of material can help delineate discrete events, this typically comes at the cost of losing temporal resolution as any stratigraphic relationship between the events is lost [@bailey2007].

The `r nrow(sites)` coastal sites chosen for analysis here have a relatively limited geographical distribution in south-eastern Norway (Figure \@ref(fig:dist)A). The sites were excavated as part of four larger excavation projects that all took place within the last 15 years [@solheim2013b; @jaksland2014a; @melvold2014b; @reitan2014l; @solheim2017b]. The sites included in the analysis consist of all Mesolithic sites excavated in conjunction with the projects that have assemblages holding more than 100 artefacts. The institution responsible for these excavations was the Museum of Cultural History in Oslo. This has led to a considerable overlap in the archaeological personnel involved, and comparable excavation practices across the excavations. Furthermore, with these projects, major efforts were made to standardise how lithic artefacts were to be classified at the museum [@melvold2014c; @koxvold2017]. As a result, this should reduce the amount of artificial patterning in the data incurred by discrepancies in the employed systems for categorisation [cf. @clark2006; @dibble2017].

The lithic data analysed is based on the classification of the site assemblages done for the original excavation reports, and consists of `r ncol(select(sites, core:flake_retouch_other))` variables representing different debitage and tool types. The artefact data have been divided into flint and non-flint materials. Flint does not outcrop naturally in southern Norway, and is only available locally as nodules that have been transported and deposited by retreating and drifting ice [e.g. @berg-hansen1999]. This means that the distribution and quality of flint has been impacted by a diverse set of climatic and geographical factors [@eigeland2015, 46]. Thus, while flint is treated as a unified category here, the variability in quality could have been substantial. Furthermore, the various non-flint raw materials that have been lumped together have quite disparate properties, where fine-grained cryptocrystalline materials are often used as a substitute or supplement to flint, while other, coarser materials are usually associated with the production of axes and other macro tools. Given this differentiated use, these raw-material properties are expected to be reflected in the retained debitage and tool categories. An important benefit of combining all of the non-flint materials is that this reduces the dependency on whether or not these have been correctly and consistently categorised for the reports [cf. @frivoll2017]. Finally, while factors such as landscape changes through shoreline displacement can have led to variable raw-material availability at the analysed sites, their relatively constrained geographical distribution hopefully counteracts some non-behavioural sources of variation.

@@

The aim of the first of part of the analysis conducted here is to evaluate the degree to which the composition of the assemblages align with earlier studies that have employed more informal methods. This therefore assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviourally meaningful. However, the approach taken is also partially informed by the so-called Frison effect [@jelinek1976], which pertains to the fact that lithics studied by archaeologists can have had long and complex use-lives in which they took on a multitude of different shapes before they were ultimately discarded. Several scholars have built on this to argue that morphological variation in retouched lithics from the Palaeolithic cannot be assumed to predominantly be the result of the intention of the original knapper to reach some desired end-product, but rather that what is commonly categorised as discrete types of artefacts by archaeologists can instead in large part be related to variable degrees of modification through use and rejuvenation [e.g. @barton1991; @dibble1995; @barton2021]. Artefact categories believed not to be internally consistent and categorically exclusive have therefore been collapsed for the analysis, as their contribution as discrete analytical units could potentially be misleading.

However, for the most part we lack even a most basic understanding of what any individual lithic object in an assemblage has been used for [@dibble2017]. For example, a vast amount of artefacts defined as debitage are likely to have fulfilled the function of tools, and both debitage and formal tool types could have had various different purposes and had a multitude of shapes throughout their use-life. While use-wear analysis could potentially offer a way to identify what artefacts were used for towards the end of their use-life, these kinds of analyses are extremely time-consuming and are therefore typically only conducted on a smaller number of artefacts that have already been selected for analysis based on their shape [e.g. @solheim2018]. Thus, while these analyses can potentially get at in-group variation pertaining to the end-state of a group of artefacts, they do not tell us whether or not their classification as a unified group is meaningful in the first place [@dibble2017]. As a consequence, the second part of the analysis employs a suite of measures developed for the classification of lithic assemblages with these inferential limitations in mind [@barton2011; @clark2017, and below]. The logic behind these measures are founded on an understanding of technology as being organised along a continuum ranging between curated and expedient [@binford1973; @binford1977; @binford1979]. An expedient technological organisation pertains to the situational production of tools to meet immediate needs, with little investment of time and resources in modification and rejuvination, resulting in high rates of tool replacement. Curated technological organisation, on the other hand, has been related to manufacture and maintenance of tools in anticipation of future use, the transport of these artefacts between places of use, and the modification and rejuvenation of artefacts for different and changing situations.

However, following not least from the ambiguous definition first put forward by @binford1973, the theoretical definition of curation, its archaeological correlates, and behavioural implications have been widely discussed and disputed [e.g. @bamforth1986; @nash1996; @shott1996; @surovell2009, 9--13]. Still, that the distinction can offer a useful analytical point of departure if clearly and explicitly operationalised seems more or less agreed upon, and some dimensions of the concept are generally accepted. For example, although precisely how it is measured may vary, the empirical correspondent to a curated technological organisation is typically defined by high degrees of retouch, as this is commonly seen as a means of realising the potential utility of a tool---or extending its use-life---by the repeated rejuvenation and modification of edges [e.g. @bamforth1986; @dibble1995; @shott2005].

One concrete operationalisation of the terms has been forwarded by @barton1998 and collegues [e.g. @barton2013; @barton1999; @riel-salvatore2004; @riel-salvatore2007; @barton2011; @barton2014; @clark2017; @villaverde1998], who through a series of studies have shown that the relationship between volumetric density of lithics and relative frequency of retouched artefacts in lithic assemblages have a consistent negative relationship across a wide range of chronological and cultural context, ranging from Pleistocene and Holocene assemblages in Europe and Asia, to assemblages associated with both Neanderthals and modern humans [@barton2011; @riel-salvatore2008]. This relationship is taken to reflect degree of curation, and is in turn mainly to follow from the accumulated nature of land-use and mobility patterns associated with the assemblages [@barton2014]. Furthermore, the relationship between curated and expedient technological organisation has been related to the continuum defined by @binford1980 between residentially mobile foragers and logistically mobile collectors [@riel-salvatore2004; @clark2017; see also @bamforth1986; @binford1977]. Residential mobility involves the movement of entire groups between resource patches throughout the year, while logistic mobility entails the use of central base-camps that are moved less often and from where task-groups venture on targeted forays to retrieve specific resources. A higher degree of logistic as opposed to residential mobility thus involves a wider range of site types and associated mobility patterns [@binford1980].

In this model, higher degree of mobility would mean a higher dependency on the artefacts and the material people could bring with them, and dimensions such as weight, reliability, repairability, and the degree to which artefacts could be manipulated to fulfil a wide range of tasks are therefore assumed to have been factors of concern. From this it follows that the empirical expectation for short-term camps is a curated technological organisation with higher relative frequency of retouched artefacts, and a lower overall density of lithics [@clark2017]. More time spent in a single location, on the other hand, is assumed to lead to better control of raw-material availability and to allow for its accumulation. This should in turn lead to a more expedient technological organisation with reduced necessity for the conservation of lithics and extensive use of retouch. The empirical expectation for lower degree of mobility is therefore relatively high density of lithics, a low relative frequency of retouched artefacts, as well as a higher number of cores and unretouched flakes and blades. These variables and underlying logic constitute what has been termed Whole Assemblage Behavioural Indicators [WABI, @clark2017], and is the main framework adopted here.

As these measures are argued to predominantly be determined by land-use and mobility patterns, relative frequency of chips and relative frequency of non-flint material are also included in the analysis, as these measures have also been linked to mobility patterns and is of central importance in Norwegian Stone Age archaeology [e.g. @Bicho2020; @breivik2016a; @reitan2016; @kitchel2021]---the use of local non-flint material has been taken to indicate reduced mobility and increased familiarity with local surroundings [@jaksland2001, 112; @glørstad2010, 181].

@@

The exploratory approach taken here means that a wide range of combinations and transformations of variables has been explored to identify patterning in the data. While only parts of this process can sensibly be reported upon, the data and employed R programming script is freely available as a research compendium at [URL placeholder], following @marwick2018b.

The `r nrow(sites)` analysed sites have been dated by reference to relative sea-level change, typology and/or radiocarbon dates (Table 1). Date ranges for sites based on shoreline displacement and typology are taken from the original reports and follow the evaluation done by the original excavators. Where radiocarbon age determinations believed to be associated with the lithic material are available, these have been calibrated using the IntCal20 calibration curve [@reimer2020] and subjected to Bayesian modelling using OxCal v4.4.4 [@bronkramsey2009] through the oxcAAR package [@hinz2021] for R [@rcoreteam]. The only constraint imposed for the modelling of the dates was that the dates from each site are assumed to represent a related group of events through the application of the Boundary function [@bronkramsey2021]. The resulting posterior density estimates were then summed for each site. Radiocarbon data is provided in the supplementary data (see also @solheim2020a).

The first part of the analysis involves employing the method of correspondence analysis (CA), using the lithic count data as classified for the original excavation reports. As this part of the analysis partially draws on the above-mentioned Frison effect, several artefact categories have been collapsed for the CA. This for example pertains to flint tool types such as scrapers, burins, drills, knives and otherwise indeterminate artefacts with retouch. These have all been combined into the single category "small flint tools". (A full overview of the aggregated variables and their constituent parts is provided in the supplementary material). While aggregating artefact categories in this manner could potentially subsume important variation, it does also reduce the possibility that any conclusions are not simply the result of employing erroneous units of analysis.

Following the WABI and other factors associated with mobility patterns, as presented above, the variables employed in the second part of the analysis are relative frequency of secondarily worked lithics (RFSL), defined as the proportion of the assemblages constituted by retouched or ground lithics; volumetric density of lithics (VDL), defined as number of artefacts per excavated m³; relative frequency of chips, defined as the proportion of artefacts with size \< 0.1 cm; relative frequency of cores, the proportion of all artefacts classified as cores in the original reports; relative frequency of blanks, here defined as the proportion of all artefacts classified as flakes, blades, micro-blades or fragments; and finally relative frequency of non-flint material. Following @Bicho2020, the analysis is done using principal components analysis (PCA), leading to a shift in focus from the relative composition emphasised by the CA, to having more weight placed on patterning in the most abundant occurrences [@baxter1994, 71--77].

A note should also be made on the fact that a few variables that are sometimes invoked for the classification of sites in terms of associated mobility patterns are omitted here [e.g. @Bicho2020; @breivik2016a]. For the assemblage data itself this especially pertains to diversity in tool-types [@canessa2021], which has been omitted in light of the above-metioned Frison effect. Number of features on the sites has also been disregarded as taphonomic loss is likely to have led to a chronological bias in their preservation. Similarly, the number of activity areas, effectively number of artefact clusters, however defined, has also been disregarded. This follows most notably from the fact that the impact of post-depositional processes at Stone Age sites in Norway is arguably understudied [@jørgensen2017]. This pertains for example to bio-turbation in the form of three-throws, which can have a detrimental effect on the original distribution of artefacts, and which can be expected to have impacted several of the sites treated here [@jørgensen2017; @darmark2018e].
